---
title: "Class 8: Breast Cancer Analysis Mini Project"
author: "Berne Chu (A18608434)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

## Data import

Data was downloaded from the class website as a CSV file.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

The first column `idgnosis` is the expert opinion on the sample(i.e. patient FNA).

```{r}
head(wisc.df$diagnosis)
```
Remove the diagnosis from data for subsequent analysis

```{r}
wisc.data <- wisc.df[,-1]
dim(wisc.data)
```

Store the diagnosis as a vector for use later when we compare our results to those from experts in the field

```{r}
diagnosis <- factor(wisc.df$diagnosis)
```

> Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/patients in the dataset

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length(grep("_mean",colnames(wisc.data)))
```

## Principal Component Analysis (PCA)

The `prcomp()` function to do PCA has a `scale=FALSE` default. In general we nearly always want to set this to TRUE so our analysis is not dominated by columns/variables in our dataset that have high standard deviation and mean when compared to others just because the units of measurement are on different units/scales.

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
summary(wisc.pr)
```
The first principal component (PC1) captures 44.3% of the original variance in the data.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 principal components.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 principal components.

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Repeat for components 1 and 3
cols <- ifelse(wisc.df$diagnosis == "M", "red", "black")

plot(wisc.pr$x[, c(1, 3) ], col = cols, 
     xlab = "PC1", ylab = "PC3")
```

The main PC result figure is called a "score plot" on "OC plot" or "ordination plot" ...

```{r}
library(ggplot2)
head(wisc.pr$x)
ggplot(wisc.pr$x)+
  aes(PC1, PC2, col=diagnosis)+
  geom_point()
```

## PCA Scree-plot

A plot of how much variance each PC captures. We can get this from `wisc.pr$sdev` from the output of `summary(wisc.pr)`

```{r}
var.tbl <- summary(wisc.pr)
```

```{r}
var = c(1,var.tbl$importance[2,])
cum.var <- c(0,var.tbl$importance[3,])

plot(var, typ="b", 
     ylab="Percent Variance Captured", 
     xlab="PC Number")
```

```{r}
plot(cum.var*100, typ="o", 
     ylab="Cumulative Percent Variance Captured", 
     xlab="PC Number",
     ylim=c(0,100))
points(var*100, col="purple", typ="o")
```

## Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean","PC1"]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

We need 5 PCs to capture more than 80% variance 

```{r}
summary(wisc.pr)
```

## Hierarchical clustering

Just clustering the original data is not very informative of helpful.

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
```

View the clustering dendrogram result

```{r}
plot(wisc.hclust)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Combining methods (PCA and Clustering)

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

Clustering the original data was not very productive. The  PCA results looked promising. Here we combine these methods by clustering from our PCA results. In other words "clustering in PC sapce"...

```{r}
dist.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist.pc, method="ward.D2")
```

View the tree

```{r}
plot(wisc.pr.hclust)
abline(h=70, col="red", lty=2)
```

To get our clustering membership vector (i.e. our main clustering result) we "cut" the three at a desired height or to yield a desired number of "k" groups.

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

How does this clustering grps compare to the expert diagnosis

```{r}
table(grps,diagnosis)
```

Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

By cutting into 2 clusters.

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The method that give my favorite result is `method="ward.D2"` because it maps better to class structure for this dataset compared to the other methods.

> Q14. How well does k-means separate the two diagnoses? How does it compare to your hclust results? (OPTIONAL)

> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
# Compare to actual diagnoses
grps4 <- cutree(wisc.pr.hclust, k = 4)
table(grps4, diagnosis)
```

It separates the two diagnoses pretty well with cluster 1 and 2 being mainly M and cluster 3 and 4 being mainly B.

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
wisc.km <- kmeans(data.scaled, centers = 2)
table(wisc.km$cluster, diagnosis)
```

Overall both clustering models did okay in separating the diagnosis with some mixed results still.

> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

The model with the best specificity was the K-means clustering on the original scaled data and the model with the best sensitivity was the PCA-based hierarchical clustering.

## 7. Prediction

We can use our PCA model for prediction with new input patient samples.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=relevel(as.factor(grps),2))
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 should be prioritized for follow-up because patient 2 in in the M cluster.
