---
title: "class 17: part 2"
author: "Berne Chu (A18608434)"
format: pdf
---

## Population Scale Analysis

> Q13.Read this file into R and determine the sample size for each genotype and 
their corresponding median expression levels for each of these genotypes.

How many sampels do we have?

```{r}
expr <- read.table("expressiondata.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Lets make a boxplot

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=T)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from 
the relative expression value between A/A and G/G displayed in this plot? Does 
the SNP effect the expression of ORMDL3?

G/G at this location results in reduced expression of this gene; A/A and A/G 
have higer expression than G/G. The SNP does influences the expression in an 
additive manner; each copy of the G allele reduces ORMDL3 expression. 
Therefore, the SNP does affect expression of ORMDL3, with G/G individuals 
showing reduced expression of the gene.